\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-10-07 chapters/patients.tex
funding

\chapter{Case studies}
\label{sec:patients}
\section{Introduction}
\todo{rename patient to bci user and participant}
\todomvh{separate terminology for sspgi}
\todo{consistently: gaze impairment, eye motor impairment or oculomotor impairment}

\todo{A systematic review of research on augmentative and alternative communication brain-computer interface systems for individuals with disabilities}

\Acp{bci} for communication assistive technology~\cite{Millan2010}
target individuals with \ac{sspi}.

visual bci

Yet, there is a large comorbidity between \ac{sspi}, and eye motor impairment~\cite{FriedOken2020}.
Impairments such as nystagmus (uncontrolled eye movements), diplopia (double
vision), and ophthalmoplegia (eye paralysis) can significantly hinder
the ability to use visual BCIs. These impairments make it difficult for
individuals to focus on or track visual stimuli accurately, reducing their
performance with BCIs that rely on visual cues~\cite{McCane2014,FriedOken2020,Pasqualotto2015}.
Unfortunately, it is again for this group that eye tracking solutions also
perform poorly, making them more reliant on potential developments in \ac{bci}
that do not rely on eye gaze.

Eye motor impairments are presumed to reduce performance in operating visual
oddball \ac{bci}s (see section\todo{ref section} for an overview), since users
cannot comfortably redirect their gaze at the desired target,
i.e., perform overt \ac{vsa}.
This is usually circumvented by designing gaze-independent~
\acp{bci}~\cite{Riccio2012}.
These interfaces either avoid visual stimulation, or exploit some form of
covert \ac{vsa}, where the gaze and \ac{vsa} do not coincide.

Several studies with visual oddball \acp{bci} show that performance drops when not fixating the intended
target~\cite{Brunner2010, Treder2010, RonAngevin2019}, necessitating
gaze-independent solutions.
These studies build on the assumption that patients with eye motor impairment
would feel comfortable operating an interface in pure covert \ac{vsa} with
central fixation.
One could argue that a \ac{bci} that is only verified to work when central
fixation is maintained, could also be considered gaze-dependent.
This does not account for the residual eye motor capabilities of most people
with \ac{sspi}, the (dis)comfort they experience while
performing gaze fixation and other confounding factors resulting from their eye
motility.

@@

\textcite{Severens2014} evaluated the visual Hex-o-Spell~\cite{Treder2010} on 5
participants with \ac{als} and showed that this visual oddball interface optimized
for gaze-independence can outperform a tactile \ac{bci}.
While this speaks to the power of visual paradigms even in groups that are
expected to have eye motor impairment, they did not verify the gaze direction
of participants during the experiment.
It was suggested that patients were performing overtly.
Participants with \ac{als} also had a substantially lower accuracy than healthy
controls (58\% vs 88\%).

Our previous study, presented in chapter \todo{ref chapter} also used the
visual Hex-o-Spell interface~\cite{VanDenKerchove2024}.
This work partially accounted for the idea that gaze impaired patients might
not fully rely on central gaze fixation and evaluates settings that are not
strictly dependent on this.
We showed gaze-independent performance can be improved
improved in healthy subjects using a suited decoding strategy.
that accounts for latency jitter in covert \ac{vsa} responses improves
Yet, there is a strong need for verification of these result in an applied
setting with people with \ac{sspi}.

\todo{discuss \cite{FriedOken2020}}

Eventually, one of the end goals of this research line is to develop
gaze-independent \ac{bci} for peopl that are fully locked-in and have no option left than to use a \ac{bci}.
However, this group is very small and it is often a challenge to recruit them
into a study and perform experiments with them~\cite{Wolpaw2006}.
Patients with less severe paralysis or in less progressed disease stages that struggle with
eye-tracking technology could also benefit from
solutions tailored to their specific situation.
Therefore, we aim to apply the concepts from earlier work an literature to
people with \ac{sspi} and various degrees motor impairment in a visuall oddball
\ac{bci}.
The objectives of this case study are as follows:
\begin{enumerate*}
  \item Explore capabilities and experienced comfort of eye motor impaired
    patients,
  \item evaluate the performance of a gaze-independent visual \ac{bci} in patients
  \item verify if this performance can be improved with a suitable decoding
    strategy.
\end{enumerate*}

\section{Materials \& methods}
\subsection{Patient recruitment}
Patients were recruited from across the Neuromuscular Reference center at
University Hospital Leuven (Leuven, Belgium), TRAINM Neuro Rehab Clinics
(Antwerp, Belgium), the Neurorehabilitation Unit at University Hospital Lille
(Lille, France) and a specialized care home (France).
Experiments were performed under the supervision of their treating physician.
Patients were recruited based on the following criteria.
To qualify for inclusion, patients must
\begin{enumerate}
	\item be at least 18 years old and no older than 60
	years,
  \item belong to class 2 or 3 according the \ac{bci}	patient selection criteria
    presented by~\textcite{Wolpaw2006},\label{item:patients/inclusion/wolpaw}
  \item have limitations to the extent or comfort of their eye motor control\label{item:patients/inclusion/oculomotor}
\end{enumerate}
Patients were excluded if they
\begin{enumerate}
  \item have a diagnosis of a major medical condition, including any major
    neurological or psychiatric disorder other than those of interest based on
    inclusion criteria~\ref{item:patients/inclusion/wolpaw},
    and~\ref{item:patients/inclusion/oculomotor}\label{item:patients/exclusion/medical}
  \item have a predisposition to or have a history of any kind of epileptic seizures,
    including photosensitive epilepsy,\label{item:patients/exclusion/epilepsy}
  \item have a severe loss in vision or hearing, that would significantly impair
        participation in the experiment,\label{item:patients/exclusion/vision}
  \item are currently using specific psychoactive medications or substances that could affect the out-
        come.\label{item:patients/exclusion/cognitive}
  \item be able to understand the experiment instructions and cooperate,
  \item have any other limitations preventing them from performing the given task.
\end{enumerate}

In total, 11 patients were contacted, of which 1 \ac{ms} patient was excluded based on
criterion~\ref{item:patients/exclusion/vision}, 1 \ac{tbi} patient on
both~\ref{item:patients/exclusion/epilepsy}
and~\ref{item:patients/exclusion/cognitive}, and one stroke patient based
on~\ref{item:patients/exclusion/medical}
One further stroke patient was excluded due to technical
difficulties during the experimental session.
Vision was assessed using a LogMAR chart~\cite{Bailey1976}.

Ultimately, 7 patients were retained.
Of these, one patient was diagnosed from bulbar-onset \ac{als}.
\ac{als} is a neurodegenerative disease affecting the motor neurons, leading to
progressive loss of motor function.
This initially results in general weakness and loss of muscle tone, but
eventually results in full body paralysis.\todo{cite}
Although speech and especially eye movements are usually preserved until the
later stages of the disease progression, the bulbar-onset variant is
characterized by an early loss of speech and an increased involvement of eye
motor symptoms~\cite{Guo2022}\todo{which?}.
Furthermore, one of the goals of \ac{bci} is to support \ac{als} patients whose
life span has been extended with life support with and assistive technology to
ensure quality of life.
In these very progressed stages of \ac{als}, eye movement will eventually also
be affected~\cite{Hayashi1991}.

Three other patients were diagnosed with \ac{fa}, a neurodegenerative
disease affecting the
spinal cord, peripheral nervous system and cerebellum.
\ac{fa} results in an impairing loss of muscle coordination.
In progressed stages, the disease can present with nystagmus, saccadic
intrusions and gaze dysmetria~\cite{Cook2017}.

The final three patients were stroke patients.
Brain stem and cerebellar stroke can often lead to forms of \ac{lis}, and
various eye motor disorders are common depending on the exact etiology
~\cite{Bogousslavsky1987, Moncayo2009}.
Stroke can lead to some of the more severe impairments like ophthalmoplegia or
partial ophthalmoplegia immediately from onset.

Table~\ref{tab:patients/patients} lists the included patients and their
diagnoses.

\todo{check Usability and Workload of Access Technology for People With Severe
Motor Impairment: A Comparison of Brain-Computer Interfacing and Eye Tracking}

\begin{table}[t]
  \centering
  \footnotesize
  \input{chapters/patients_table.tex}
  \caption[Presentation of included patients including their diagnosis and
  capabilities.]{Presentation of included patients including their diagnosis and
  capabilities.
  (Trach.: patient underwent a tracheostomy, Cls.: classification according
  to~\textcite{Wolpaw2006}).
  }
  \label{tab:patients/patients}
\end{table}
\todo{Replace with classification according to 40. Kübler A, Birbaumer N.
Brain-computer interfaces and communication in paralysis: extinction of goal
directed thinking in completely paralysed patients? Clin Neurophysiol.
2008;119:2658-2666.}
\todo{include year of diagnosis}

@@

As an objective metric, we implemented and performed the automated NeuroEye eye movement
test proposed by~\textcite{Hassan2022} using calibration-free eye tracking to
check if it revealed any further eye motor abnormalities.
This was not the case.
This test is based on calibration-free eye tracking.
\todo{say which part we used and which part not}
\todo{motivation and text and references Juliette}
\todo{Limited in what it can tell us (check paper)}
\todo{include eye test document juliette}

Table~\ref{tab:patients/eye} details their eye motor impairments and vision of
included patients.
All participants reported some degree of fatigue or discomfort when fixating.
Patient PA1 had the mildest impairment, only reporting fatigue when fixating
for prolonged times.
The \ac{fa} patients were mostly affected by eye tremors and impaired pursuit.
Eye motor function of participants PC2, PC3, and PC3 were most severly affected.
Participant PC2 was only able to look up and down and had a deviation in the
left eye causing diplopia, but this was corrected by a prism glass.
Participant PC3 only retained partial motility of the right eye, while the left eye was permanently closed.
Participant PC4 had one deviated eye with a corneal abcess affecting the motility
and vision in the right eye, and suffered reduced motility in the left.
\todo{go through notes again}

\begin{table}[t]
  \centering
  \footnotesize
  \input{chapters/patients_table_eye.tex}
  \caption[Visual skills of the included participants.]{%
  Visual skills of the included participants.
  Visual~\ac{bci} skills~\cite{FriedOken2020}
  was assessed with a combination of self-reported issues
  by the subject and the NeuroEye~\cite{Hassan2022} test. \BigCross\
  impaired, \BigDiamondshape\ severely impaired. Visual acuity in logMAR, lower
  is better.}
  \label{tab:patients/eye}
\end{table}

@@

3 different \ac{vsa} settings were explored
In overt \ac{vsa}, the participant was instructed to fixate the cued target, or
try to the maximum extent of their visual skill, even if experiencing slight
discomfort.
In covert \ac{vsa}, the participant was instructed to fixate the center of the
screen, also to the extent of their ability.
An additional \emph{free \ac{vsa}} condition was introduced.
Here, the patient was instructed to perform the task as they deemed most
comfortable.
This allows us to investigate the user's natural way of operating the \ac{bci}
given their individual set of visual skills.
If the participant was not fully paralyzed, they were instructed not to move the head.
The cued \emph{split attention} setting proposed
by~\textcite{VanDenKerchove2024} is not studied here, since we are interested
in natural \ac{vsa} operation settings for gaze-impaired individuals.

To make the interface suitable for use by individuals with
\ac{sspi}~\cite{FriedOken2020}, the
number of blocks was decreased to 6 per \ac{vsa} setting.
\Ac{isi} was increased to\todo{number} to decrease task difficulty.
The experiment also started with a training block in each condition, where the
participant was instructed with feedback on their performance to make sure they
understood and were able to perform the task.

\begin{figure}
  \begin{subfigure}{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/patients/PD01a-obfuscated.jpg}
    \caption{A patient with the stimulation and recording setup.}
    \label{fig:patients/photos/side}
  \end{subfigure}\hfill%
  \begin{subfigure}{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/patients/PD01b-obfuscated.jpg}
    \caption{caption}
    \label{fig:patients/photos/front}
  \end{subfigure}
  \caption{The recording setup and stimulation interface.}
  \label{fig:patients/photos}
\end{figure}
\todo{caption}
\todo{include stimulation interface}

@@

During the recording session, patients were positioned in their wheelchair in front of a table.
Stimuli were presented on a laptop \todo{model, make} placed at \todo{distance}
A Cedrus StimTracker \todo{make} ensured synchronization of stimuli with the
recorded \ac{eeg}.
Eye tracking throughout was performed using the Tobii~\todo{model, make}
portable eye tracker placed at the bottom of the laptop screen.

@@

First, we aimed to shed more light on the actual capabilities of gaze impaired
individuals with respect to performing overt \ac{vsa} and central gaze
fixation, as well as investigate how relevant these two settings are when the
gaze is not cued.
Figure~\ref{fig:patients/gaze} maps gaze position relative to the stimuli
across conditions.
These results should be interpreted with care, since the eye tracker to some
degree relies on functioning eye motility.
The participant's position
relative due the eye tracker might have shifted throughout the experimental
session despite our best efforts, e.g.\ because they needed aspiration of their
tracheostomy.

\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/patients/fig_gaze.png}
}{%
  Distribution of the recorded gaze position.
}{%
  Distribution of the recorded gaze position during the experimental session in the three \ac{vsa}
  conditions.
  Subjects PB2 and PC4 preferred covert \ac{bci} operation, with PB2 resting gaze
  near the middle of the screen, and PC4 near the bottom.
}{%
  fig:patients/gaze
}
\todo{explaine crosses}

@@

\begin{figure}[t]
  \input{figures/patients/fig_decode.pgf}
  \caption[%
    Decoding performance in different \ac{vsa} settings.
  ]{%
    Decoding performance in different \ac{vsa} settings reported as
    single-trial \ac{rocauc}.
    Free \ac{vsa} is generally on par with overt \ac{vsa}.
    Performance in covert \ac{vsa} with central gaze fixation is lower, but can
    be improved with the \ac{wcble} decoder.
    95\% confidence intervals were calculated using 1000 bootstrapping
    repetitions.
  }
  \label{fig:patients/decode}
\end{figure}
\todo{rename subjects}
\todo{check if percentages or 0-1 accuracy and roc-auc}

In overt \ac{vsa}\todomvh{specify setting when talking about the setting}, evaluated decoders performed about equal on average
(\ac{wcble} 75.58\%, XDAWNCov+TS+LDA 74.24\%, \ac{tlda} 75.99\%).
In covert \ac{vsa} with cued central gaze fixation, performance deteriorates
but \ac{wcble} can significantly improve performance over its base classifier
\ac{tlda} in this condition
(\ac{wcble} 62.49\%, XDAWNCov+TS+LDA 59.42\%, \ac{tlda} 59.05\%).
Decoding performance for this task was at chance level for participants PB4 and
PC3.

However, \Ac{wcble} did not improve \ac{tlda} in free \ac{vsa} free \ac{vsa}, but
XDAWNCov+TS+LDA performance was slightly lower here (although not
significantly).
(\ac{wcble} 74.15\%, XDAWNCov+TS+LDA 71.88\%, \ac{tlda} 74.27\%).
More interestingly, we notice that performances of the decoders in free
\ac{vsa} are close to those in overt \ac{vsa}.
A substantial decrease in performance from overt to free is present for
subjects PC3 and PC4.
For PB2, who also relied on covert \ac{vsa} during the uncued free \ac{vsa}
setting, there was no decrease in performance.

@@

Due to the heterogenous nature of the participant's conditions, it is hard to support
claims about significance of presented results.\todomvh{
weird formulation
hard to draw general conclusions}
This study should therefore be seen as a collection of case studies,
highlighting different obstacles encountered in developing gaze-independent
visual oddball \acp{bci} for individuals with \ac{sspi} and gaze impairment.
Nevertheless, we would like to highlight some general aspects that might be
of interest for the further development of this class of \acp{bci}.

True gaze-indepedent visual \acp{bci} should not rely on gaze fixation.\todomvh{any reference/evidence in support of this statement?}
Hence our analysis centers around the free \ac{vsa} condition.
Eye tracking results presented in section~\ref{sec:patients/outcomes/gaze}
confirm our assumption that voluntary covert \ac{vsa} can
occur in individuals with \ac{sspi} and gaze impairment.
We have also confirmed part of the results from~\textcite{VanDenKerchove2024} presented in
chapter\todo{ref} that state that decoding of covert \ac{vsa} with central gaze
fixation can be improved by accounting for latency jitter, and showed that this
holds for individuals with \ac{sspi} and gaze impairment.

@@

The seamingly stable performance across overt and free \ac{vsa} could be
misinterpreted as an indication that the Hex-o-Spell \ac{bci} already works
well for patients with \ac{sspi} and gaze impairment, and no optimization is
needed.
However, we assume that overt \ac{vsa} performance was also decreased in some
subjects or for some blocks if the participant was not able to comfortably
perform the task.
Nevertheless, the large difference between free \ac{vsa} and covert \ac{vsa}
with central gaze fixation, is food for thought about the applicability of
solutions developed with central fixation in mind.

@@

First and foremost, this study works with a limited sample of patients which
does not represent the full spectrum of individuals with \ac{sspi} and their
oculomotor impairment.
\Ac{fa} patients met the inclusion criteria, but they are usually not
considered as one of the typical interest groups of \ac{bci} communication assistive
technology, partly due to the rarity, and partly due to the disease
progression.
It would be most interesting to verify these results with individuals with
\ac{lis} and no eye movement capability at all.

Another limiting factor are the difficulties in correctly interpreting eye
tracker results in studies with eye motor impaired patients.
If eye tracking at all is possible, it is not guaranteed that the user is able
to perform the calibration procedure successfully.
Further experiments should be carried out with a stationary eye tracker with
more advanced capabilities, although systems using a head fixator or headrest
should be avoided.
This is not practical when working with
wheelchair-bound individuals who might have undergone a tracheostomy and might
suffer from spasticity.

In this study, user comfort in the different conditions was not objectively
measured.
Rather, it was assumed that patients operated most comfortably in the free
\ac{vsa} condition.
In order to properly contextualize performance results, they should be coupled
with a measure of user comfort and percieved effort, like the NASA Task Load
Index~\cite{Hart2006}.
Performance might after all be traded off for user comfort.
Eye motor disability can also have been assessed more objectively, using \todo{

>>>>

\chapter{Case studies}
\label{sec:patients}
\section{Introduction}

\Acp{bci} for communication assistive technology~\cite{Millan2010}
target individuals with \ac{sspi}~\cite{Peters2022}.

\todo{visual bci}

Yet, there is a large comorbidity between \ac{sspi}, and eye motor impairment~\cite{FriedOken2020}.
Impairments such as nystagmus (uncontrolled eye movements), diplopia (double
vision), and ophthalmoplegia (eye paralysis) can significantly hinder
the ability to use visual BCIs. These impairments make it difficult for
\ac{bci} users to focus on or track visual stimuli accurately, reducing their
performance with BCIs that rely on visual cues~\cite{McCane2014,FriedOken2020,Pasqualotto2015}.
Unfortunately, it is again for this group that eye tracking solutions also
perform poorly, making them more reliant on potential developments in \ac{bci}
that do not rely on eye gaze.

Eye motor impairments are presumed to reduce performance in operating visual
oddball \ac{bci}s (see \autoref{sec:gaze-independence} for an overview), since users
cannot comfortably redirect their gaze at the desired target,
i.e., perform overt \ac{vsa}.
This is usually circumvented by designing gaze-independent~
\acp{bci}~\cite{Riccio2012}.
These interfaces either avoid visual stimulation, or exploit some form of
covert \ac{vsa}, where the gaze and \ac{vsa} do not coincide.

Several studies with visual oddball \acp{bci} show that performance drops when not fixating the intended
target~\cite{Brunner2010, Treder2010, RonAngevin2019}, necessitating
gaze-independent solutions.
These studies build on the assumption that \ac{bci} users with \ac{sspgi}
would feel comfortable operating an interface in pure covert \ac{vsa} with
central fixation.
One could argue that a \ac{bci} that is only verified to work when central
fixation is maintained, could also be considered gaze-dependent.
This does not account for the residual eye motor capabilities of most people
with \ac{sspi}, the (dis)comfort they experience while
performing gaze fixation and other confounding factors resulting from their eye
motility.

@@

\textcite{Severens2014} evaluated the visual Hex-o-Spell~\cite{Treder2010} on 5
participants with \ac{als} and showed that this visual oddball interface optimized
for gaze-independence can outperform a tactile \ac{bci}.
While this speaks to the power of visual paradigms even in groups that are
expected to have eye motor impairment, they did not verify the gaze direction
of participants during the experiment.
It was suggested that participants were performing overtly.
Participants with \ac{als} also had a substantially lower accuracy than healthy
controls (58\% vs 88\%).

Our previous study, presented in chapter \autoref{sec:covert-align} also used the
visual Hex-o-Spell interface~\cite{VanDenKerchove2024}.
This work partially accounted for the idea that \ac{bci} users with \ac{sspgi} might
not fully rely on central gaze fixation and evaluates settings that are not
strictly dependent on this.
We showed gaze-independent performance can be improved
improved in healthy subjects using a suited decoding strategy.
that accounts for latency jitter in covert \ac{vsa} responses improves
Yet, there is a strong need for verification of these result in an applied
setting with people with \ac{sspi}.

%\todo{discuss \cite{FriedOken2020}}

Eventually, one of the end goals of this research line is to develop
gaze-independent \ac{bci} for peopl that are fully locked-in and have no option left than to use a \ac{bci}.
However, this group is very small and it is often a challenge to recruit them
into a study and perform experiments with them~\cite{Wolpaw2006}.
Patients with less severe paralysis or in less progressed disease stages that struggle with
eye-tracking technology could also benefit from
solutions tailored to their specific situation.
Therefore, we aim to apply the concepts from earlier work an literature to
people with \ac{sspi} and various degrees motor impairment in a visuall oddball
\ac{bci}.
The objectives of this case study are as follows:
\begin{enumerate*}
  \item Explore capabilities and experienced comfort of individuals with
    \ac{sspgi}, when operating a visual \ac{bci},
  \item evaluate the performance of a gaze-independent visual \ac{bci} for this
    group,
  \item verify if this performance can be improved with a suitable decoding
    strategy.
\end{enumerate*}

\section{Materials \& methods}
\subsection{Patient recruitment}
Patients were recruited from across the Neuromuscular Reference center at
University Hospital Leuven (Leuven, Belgium), TRAINM Neuro Rehab Clinics
(Antwerp, Belgium), the Neurorehabilitation Unit at University Hospital Lille
(Lille, France) and a specialized care home (France).
Experiments were performed under the supervision of their treating physician.
Patients were recruited based on the following criteria.
To qualify for inclusion, participants must
\begin{enumerate}
	\item be at least 18 years old and no older than 60
	years,
  \item belong to class 2 or 3 according the \ac{bci}	user selection criteria
    presented by~\textcite{Wolpaw2006},\label{item:patients/inclusion/wolpaw}
  \item have limitations to the extent or comfort of their eye motor control\label{item:patients/inclusion/oculomotor}
\end{enumerate}
Patients were excluded if they
\begin{enumerate}
  \item have a diagnosis of a major medical condition, including any major
    neurological or psychiatric disorder other than those of interest based on
    inclusion criteria~\ref{item:patients/inclusion/wolpaw},
    and~\ref{item:patients/inclusion/oculomotor}\label{item:patients/exclusion/medical}
  \item have a predisposition to or have a history of any kind of epileptic seizures,
    including photosensitive epilepsy,\label{item:patients/exclusion/epilepsy}
  \item have a severe loss in vision or hearing, that would significantly impair
        participation in the experiment,\label{item:patients/exclusion/vision}
  \item are currently using specific psychoactive medications or substances that could affect the out-
        come.\label{item:patients/exclusion/cognitive}
  \item be able to understand the experiment instructions and cooperate,
  \item have any other limitations preventing them from performing the given task.
\end{enumerate}

In total, 11 individuals were contacted, of which one person with \ac{ms} was excluded based on
criterion~\ref{item:patients/exclusion/vision}, one person recovering from
\ac{tbi} on
both~\ref{item:patients/exclusion/epilepsy}
and~\ref{item:patients/exclusion/cognitive}, and one person recovering from
stroke based
on~\ref{item:patients/exclusion/medical}
One further person recovering from a stroke was excluded due to technical
difficulties during the experimental session.
Vision was assessed using a LogMAR chart~\cite{Bailey1976}.

Ultimately, 7 participants were retained.
Of these, one participant was diagnosed from bulbar-onset \ac{als}.
\ac{als} is a neurodegenerative disease affecting the motor neurons, leading to
progressive loss of motor function.
This initially results in general weakness and loss of muscle tone, but
eventually results in full body paralysis.
Although speech and especially eye movements are usually preserved until the
later stages of the disease progression, the bulbar-onset variant is
characterized by an early loss of speech and an increased involvement of eye
motor symptoms~\cite{Guo2022}.
Furthermore, one of the goals of \ac{bci} is to support individuals with \ac{als}
whose life span has been extended with life support with and assistive technology to
ensure quality of life.
In these very progressed stages of \ac{als}, eye movement will eventually also
be affected~\cite{Hayashi1991}.

Three other participants were diagnosed with \ac{fa}, a neurodegenerative
disease affecting the
spinal cord, peripheral nervous system and cerebellum.
\ac{fa} results in an impairing loss of muscle coordination.
In progressed stages, the disease can present with nystagmus, saccadic
intrusions and gaze dysmetria~\cite{Cook2017}.

The final three participants were recovering from stroke.
Brain stem and cerebellar stroke can often lead to forms of \ac{lis}, and
various eye motor disorders are common depending on the exact etiology
~\cite{Bogousslavsky1987, Moncayo2009}.
Stroke can lead to some of the more severe impairments like ophthalmoplegia or
partial ophthalmoplegia immediately from onset.

Table~\ref{tab:patients/patients} lists the included participants and their
diagnoses.

\newcommand{\fnkuebler}{%
  \footnote{
``With minor degree of impairment,
we refer to patients who had only slightly impaired limb movement
and normal speech. Under the category moderate impairment, we
summarized those patients with restricted limb movement (wheelchair-bound) and
unaffected speech or intact limb movement without speech (such as the bulbar form of ALS which first affects speech
and swallowing). Patients who were almost tetraplegic with restricted speech were considered major impaired. Categories four
and five were the LIS and the CLIS, respectively."~\cite{Kuebler2008}
  }
}
\begin{table}[t]
  \centering
  \makebox[\textwidth][c]{%
  \footnotesize
  \input{chapters/patients_table.tex}
  }
  \caption[Included participants with their diagnosis and
  capabilities.]{Included participants with their diagnosis and
  capabilities.
  Trach.: underwent a tracheostomy, Cls. W: classification according
  to~\textcite{Wolpaw2006}, Cls. KB: classification according
  to~\textcite{Kuebler2008}\fnkuebler.
  }
  \label{tab:patients/patients}
\end{table}
\todo{Replace with classification according to 40. Kübler A, Birbaumer N.
Brain-computer interfaces and communication in paralysis: extinction of goal
directed thinking in completely paralysed patients? Clin Neurophysiol.
2008;119:2658-2666.}
\todo{include year of diagnosis}

@@

As an objective metric, we implemented and performed the automated NeuroEye eye movement
test proposed by~\textcite{Hassan2022} using calibration-free eye tracking to
check if it revealed any further eye motor abnormalities.
This was not the case.

Table~\ref{tab:patients/eye} details their eye motor impairments and vision of
included participants.
All participants reported some degree of fatigue or discomfort when fixating.
Patient PA1 had the mildest impairment, only reporting fatigue when fixating
for prolonged times.
The \ac{fa} participants were mostly affected by eye tremors and impaired pursuit.
PB2 suffered from especially severe horizontal oscillating involuntary eye
movements.
Eye motor function of participants PC2, PC3, and PC3 were most severly affected.
Participant PC2 was only able to look up and down and had a deviation in the
left eye causing diplopia, but this was corrected by a prism glass.
Participant PC3 only retained partial motility of the right eye, while the left eye was permanently closed.
Participant PC4 had one deviated eye with a corneal abcess affecting the motility
and vision in the right eye, and suffered reduced motility in the left.

\begin{table}[t]
  \centering
  \footnotesize
  \input{chapters/patients_table_eye.tex}
  \caption[Visual skills of the included participants.]{%
  Visual skills of the included participants.
  Visual~\ac{bci} skills~\cite{FriedOken2020}
  were assessed with a combination of self-reported issues
  by the subject and the NeuroEye~\cite{Hassan2022} test. \BigCross\
  impaired, \BigDiamondshape\ severely impaired.
  logMAR: lower is better.}
  \label{tab:patients/eye}
\end{table}

@@

3 different \ac{vsa} settings were explored
In the overt \ac{vsa} setting, the participant was instructed to fixate the cued target, or
try to the maximum extent of their visual skill, even if experiencing slight
discomfort.
In the covert \ac{vsa} setting, the participant was instructed to fixate the center of the
screen, also to the extent of their ability.
An additional \emph{free \ac{vsa}} setting was introduced.
Here, the participant was instructed to perform the task as they deemed most
comfortable.
This allows us to investigate the user's natural way of operating the \ac{bci}
given their individual set of visual skills.
If the participant was not fully paralyzed, they were instructed not to move the head.
The cued \emph{split attention} setting proposed
by~\textcite{VanDenKerchove2024} is not studied here, since we are interested
in natural \ac{vsa} operation settings for gaze-impaired individuals.

To make the interface suitable for use by individuals with
\ac{sspi}~\cite{FriedOken2020}, the
number of blocks was decreased to 6 per \ac{vsa} setting.
\Ac{isi} was increased to $200\pm50$ms to decrease task difficulty.
The experiment also started with a training block in each condition, where the
participant was instructed with feedback on their performance to make sure they
understood and were able to perform the task.

\begin{figure}
  \begin{minipage}{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/patients/PD01a-obfuscated.jpg}
  \end{minipage}\hfill%
  \begin{minipage}{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/patients/PD01b-obfuscated.jpg}
  \end{minipage}
  \caption{A participant with the stimulation and recording setup.}
\end{figure}

@@

During the recording session, participants were positioned in their wheelchair in front of a table.
Stimuli were presented on a Acer Predator Helios laptop with 18" screen (Acer,
Inc., Taiwan) placed at a 60cm distance.
A Cedrus StimTracker (Cedrus Corp., CA, USA) ensured synchronization of stimuli with the
recorded \ac{eeg}.
Eye tracking throughout was performed using the Tobii~\todo{model, make}
portable eye tracker placed at the bottom of the laptop screen.

@@

First, we aimed to shed more light on the actual capabilities of individuals
with \ac{sspgi} with respect to performing overt \ac{vsa} and central gaze
fixation, as well as investigate how relevant these two settings are when the
gaze is not cued.
Figure~\ref{fig:patients/gaze} maps gaze position relative to the stimuli
across conditions.
These results should be interpreted with care, since the eye tracker to some
degree relies on functioning eye motility.
The participant's position
relative due the eye tracker might have shifted throughout the experimental
session despite our best efforts, e.g.\ because they needed aspiration of their
tracheostomy.

\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/patients/fig_gaze.png}
}{%
  Distribution of the recorded gaze position.
}{%
  Distribution of the recorded gaze position during the experimental session in the three \ac{vsa}
  conditions.
  Crosshairs represent stimulus positions, with the orange ones cued during
  the given condition.
  Subjects PB2 and PC4 preferred covert \ac{bci} operation, with PB2 resting gaze
  near the middle of the screen, and PC4 near the bottom.
}{%
  fig:patients/gaze
}

@@

\begin{figure}[t]
  \input{figures/patients/fig_decode.pgf}
  \caption[%
    Decoding performance in different \ac{vsa} settings.
  ]{%
    Decoding performance in different \ac{vsa} settings reported as
    single-trial \ac{rocauc}.
    Free \ac{vsa} is generally on par with performance in the overt \ac{vsa}
    settign.
    Performance in the covert \ac{vsa} setting with central gaze fixation is lower, but can
    be improved with the \ac{wcble} decoder.
    95\% confidence intervals were calculated using 1000 bootstrapping
    repetitions.
  }
  \label{fig:patients/decode}
\end{figure}
\todo{figure margin}
\todo{check if percentages or 0-1 accuracy and roc-auc everywhere}

In the overt \ac{vsa} setting, evaluated decoders performed about equal on average
(\ac{wcble} 75.58\%, XDAWNCov+TS+LDA 74.24\%, \ac{tlda} 75.99\%).
In the covert \ac{vsa} setting with cued central gaze fixation, performance deteriorates
but \ac{wcble} can significantly improve performance over its base classifier
\ac{tlda} in this condition
(\ac{wcble} 62.49\%, XDAWNCov+TS+LDA 59.42\%, \ac{tlda} 59.05\%).
Decoding performance for this task was at chance level for participants PB4 and
PC3.

However, \Ac{wcble} did not improve \ac{tlda} in free \ac{vsa} free \ac{vsa}, but
XDAWNCov+TS+LDA performance was slightly lower here (although not
significantly).
(\ac{wcble} 74.15\%, XDAWNCov+TS+LDA 71.88\%, \ac{tlda} 74.27\%).
More interestingly, we notice that performances of the decoders in free
\ac{vsa} are close to those in the overt \ac{vsa}.
A substantial decrease in performance from the overt setting to the free
setting is present for
subjects PC3 and PC4.
For PB2, who also relied on covert \ac{vsa} during the uncued free \ac{vsa}
setting, there was no decrease in performance.

@@

Due to the heterogenous nature of the participant's conditions, it is hard to
draw general conclusions.
This study should therefore be seen as a collection of case studies,
highlighting different obstacles encountered in developing gaze-independent
visual oddball \acp{bci} for individuals with \ac{sspgi}.
Nevertheless, we would like to highlight some aspects that might be
of interest for the further development of this class of \acp{bci}.

True gaze-indepedent visual \acp{bci} should not rely on gaze fixation.
Hence our analysis centers around the free \ac{vsa} condition.
Eye tracking results presented in section~\ref{sec:patients/outcomes/gaze}
confirm our assumption that voluntary covert \ac{vsa} can
occur in individuals with \ac{sspgi}.
We have also confirmed part of the results from~\textcite{VanDenKerchove2024} presented in
chapter\todo{ref} that state that decoding of covert \ac{vsa} with central gaze
fixation can be improved by accounting for latency jitter, and showed that this
holds for individuals with \ac{sspgi}.

@@

The seamingly stable performance across overt and free \ac{vsa} could be
misinterpreted as an indication that the Hex-o-Spell \ac{bci} already works
well for individuals with \ac{sspgi}, and no optimization is
needed.
However, we assume that overt \ac{vsa} performance was also decreased in some
subjects or for some blocks if the participant was not able to comfortably
perform the task.
Nevertheless, the large difference between the free \ac{vsa} setting and the
covert \ac{vsa} setting
with central gaze fixation, is food for thought about the applicability of
solutions developed with central fixation in mind.

@@

First and foremost, this study works with a limited sample size which
does not represent the full spectrum of individuals with \ac{sspi} and
\ac{sspgi}, and their specific symptoms and skills?
Individuals with \ac{fa} met the inclusion criteria, but they are usually not
considered as one of the typical interest groups of \ac{bci} communication assistive
technology, partly due to the rarity, and partly due to the disease
progression.
It would be most interesting to verify these results with individuals with
\ac{lis} and no eye movement capability at all.

Another limiting factor are the difficulties in correctly interpreting eye
tracker results in studies with individuals with gaze impairment.
If eye tracking at all is possible, it is not guaranteed that the user is able
to perform the calibration procedure successfully.
Further experiments should be carried out with a stationary eye tracker with
more advanced capabilities, although systems using a head fixator or headrest
should be avoided.
This is not practical when working with
wheelchair-bound individuals who might have undergone a tracheostomy and might
suffer from spasticity.

In this study, user comfort in the different conditions was not objectively
measured.
Rather, it was assumed that participants operated most comfortably in the free
\ac{vsa} condition.
In order to properly contextualize performance results, they should be coupled
with a measure of user comfort and percieved effort, like the NASA Task Load
Index~\cite{Hart2006}.
Performance might after all be traded off for user comfort.
Eye motor disability can also have been assessed more objectively, using \todo{
